Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-07-02 paper/paper.tex
revision

Normally, if the continuation @konts@ is an empty list, then it means we have reached
the end of the computation (i.e., a halt); otherwise, we should return the values of 
@ae@ to its caller which is contained in the top frame of stack.

However, since we have added nondeterminism into state, an empty list of frames means
we have reached the halt of \textit{one computation path}, and we should determine the
next state indicated by @NDCont@. Thus, we have an inner pattern matching on @ndk@:

@@

\begin{lstlisting}
  def continue(nds: NDState, seen: Set[State]): Set[State] = {
    val NDState(ae, env, bstore, konts, time, ndk) = nds; val new_time = nds.tick
    konts match {
      case Nil => ndcontinue(nds, seen)
      case Frame(x, e, f_env)::konts =>
        val baddr = allocBind(x, new_time); val new_env = f_env + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        drive_step(NDState(e, new_env, new_store, konts, new_time, ndk), seen)
    }
  }
\end{lstlisting}

@@

\begin{lstlisting}
  def ndcontinue(nds: NDState, seen: Set[State]): Set[State] = {
    val NDState(ae, env, bstore, konts, time, ndk) = nds
    ndk match {
      case NDCont(Nil, _, _, _, _)::ndk =>
        drive_step(NDState(ae, env, bstore, konts, time, ndk), seen)
      case NDCont(cls, argvs, bstore, time, frames)::ndk =>
        val Clos(Lam(v, body), c_env) = cls.head
        val baddr = allocBind(v, time); val new_env = c_env + (v -> baddr)
        val new_store = bstore.update(baddr, argvs)
        val new_ndk = NDCont(cls.tail, argvs, bstore, time, frames)::ndk
        drive_step(NDState(body, new_env, new_store, frames, time, new_ndk, seen))
    }
  }
\end{lstlisting}

@@

\begin{lstlisting}
  case Let(x, App(f, ae), e) =>
    val closures = atomicEval(f, env, store)
    nd[Storable](closures, Set[VS](), { case (clos, clsnd) =>
      val Clos(Lam(v, body), c_env) = clos
      val baddr = allocBind(v, new_time); val new_cenv = c_env + (v -> baddr)
      val new_store = store.update(baddr, atomicEval(ae, env, store))
      aeval(body, new_cenv, new_store, new_time, (bdvss: Set[VS]) => {
        nd[VS](bdvss, Set[VS](), { case (vs, bdnd) =>
          val VS(vals, time, store) = vs
          val baddr = allocBind(x, time); val new_env = env + (x -> baddr)
          val new_store = store.update(baddr, vals)
          /* evaluate e with new env and store, then transfer to next body's value */
          aeval(e, new_env, new_store, time, bdnd)
        }, clsnd) /* transfer to next target closure of f */
      })
    },
    continue)
\end{lstlisting}

@@

Besides, Danvy and his collaborators observed the correspondence between evaluation contexts and continuations.
For deterministic languages, \citet{Danvy:2008:DIP:1411204.1411206} showed that reduction contexts
are defunctionalized (data type representing) continuations and reduction semantics is a structural 
operational semantics in continuation-passing style.
Not unexpectedly, the correspondence also exists for nondeterministic languages. For example,
\citet{Danvy:2001:DW:773184.773202} showed that refunctionalizing a regular expression matcher yields
its counterpart with two layer of continuations.
In addition, the operational semantics of control operators (e.g., @shift@ and @reset@) can be given by an 
abstract machine with multi-layer continuations \cite{Danvy:1990:AC:91556.91622, lmcs:2269}.

>>>>

Normally, if the continuation @konts@ is an empty list, then it means we have reached
the end of the computation (i.e., a halt); otherwise, we should return the values of 
@ae@ to its caller which is contained in the top frame of the stack.
However, since we have added a meta-continuation into the state, an empty list of frames 
means we have reached the halt of \textit{one computation path}, and we should determine the
next state indicated by @NDCont@. Thus, we have an inner pattern matching on @ndk@:

@@

\begin{lstlisting}
  def continue(nds: NDState, seen: Set[State]): Set[State] = {
    val NDState(ae, env, bstore, konts, time, ndk) = nds; val new_time = nds.tick
    konts match {
      case Nil => ndcontinue(nds, seen)
      case Frame(x, e, f_env)::konts =>
        val baddr = allocBind(x, new_time); val new_env = f_env + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        drive_step(NDState(e, new_env, new_store, konts, new_time, ndk), seen) /* normal return */
    }
  }
\end{lstlisting}

@@

\begin{lstlisting}
  def ndcontinue(nds: NDState, seen: Set[State]): Set[State] = {
    val NDState(ae, env, bstore, konts, time, ndk) = nds
    ndk match {
      case NDCont(Nil, _, _, _, _)::ndk =>
        /* transfer to the most recent fork point */
        drive_step(NDState(ae, env, bstore, konts, time, ndk), seen) 
      case NDCont(cls, argvs, bstore, time, frames)::ndk =>
        val Clos(Lam(v, body), c_env) = cls.head
        val baddr = allocBind(v, time); val new_env = c_env + (v -> baddr)
        val new_store = bstore.update(baddr, argvs)
        val new_ndk = NDCont(cls.tail, argvs, bstore, time, frames)::ndk
        /* resume the fork point with the next closure */
        drive_step(NDState(body, new_env, new_store, frames, time, new_ndk, seen))
    }
  }
\end{lstlisting}

@@

\begin{lstlisting}
  case Let(x, App(f, ae), e) =>
    val closures = atomicEval(f, env, store)
    nd[Storable](closures, Set[VS](), { case (clos, clsnd) =>
      val Clos(Lam(v, body), c_env) = clos
      val baddr = allocBind(v, new_time); val new_cenv = c_env + (v -> baddr)
      val new_store = store.update(baddr, atomicEval(ae, env, store))
      aeval(body, new_cenv, new_store, new_time, (bdvss: Set[VS]) => {
        nd[VS](bdvss, Set[VS](), { case (vs, bdnd) =>
          val VS(vals, time, store) = vs
          val baddr = allocBind(x, time); val new_env = env + (x -> baddr)
          val new_store = store.update(baddr, vals)
          /* evaluate e with new env and store, then transfer to the next body's value */
          aeval(e, new_env, new_store, time, bdnd)
        }, clsnd) /* transfer to the next target closure of f */
      })
    },
    continue)
\end{lstlisting}

@@

Besides, Danvy and his collaborators observed the correspondence between evaluation contexts and continuations.
For deterministic languages, \citet{Danvy:2008:DIP:1411204.1411206} showed that reduction contexts
are defunctionalized (data type representing) continuations and reduction semantics is a structural 
operational semantics in continuation-passing style.
Not unexpectedly, the correspondence also exists for nondeterministic languages. For example,
\citet{Danvy:2001:DW:773184.773202} observed that refunctionalizing a regular expression matcher 
with two stacks yields its counterpart with two layer of continuations.
In addition, the operational semantics of control operators (e.g., @shift@ and @reset@) can be given by an 
abstract machine with multi-layer continuations \cite{Danvy:1990:AC:91556.91622, lmcs:2269}.
